---
---

## Exercises

### Exercise 1

Use the quadratic formula to find $$x$$ that satisfies the
equation $$1.5 x^2 + 0.3 x - 2.9 = 0$$.

$$
\frac{-0.3 \pm \sqrt{0.3^2 - 4 \times 1.5 \times {-2.9}}}{2 \times 1.5}
$$

### Exercise 2

By default, all character data is read in to a data.frame as factors. Use the
`read.csv()` argument `stringsAsFactors` to suppress this behavior, then
subsequently modify the `STATE` column in `storm` to make it a factor. Remember
that columns of a `data.frame` are identified to the R interpreter with the `$`
operator, e.g. `storm$STATE`.

### Exercise 3

Use the `typeof` function to inspect the data type of `income`, and do the same
for another variable to which you assign a `list` of numbers. Why are they
different? Use `c` to combine `income` with the new variable you just created
and inspect the result with `typeof`. Does `c` always create vectors?

### Exercise 4

Create a data frame with two columns, one called "species" with four strings and
another called "abund" with four numbers. Store your data frame as a variable
called `data`.

### Exercise 5

1. Get weekdays using negative integers.
1. Get M-W-F using a vector of postitions generated by `seq()` that uses the
`by` argument (don't forget to `?seq` for help).

### Exercise 6

The keywords `else` and `if` can be combined to allow flow control among more
than two statements, as below. Expand the `first` function once again to
differentiate between `dat` provided as a `matrix` and as a `data.frame`. It's
up to you what the "first" element of a matrix should be!

```{r, eval = FALSE, title = "Exercise"}
if (...) {
  ...
} else if {
  ...
} else {
  ...
}
```

## Solutions

### Solution 1

```{r, title = "Solution"}
(-0.3 + sqrt(0.3 ^ 2 - 4 * 1.5 * -2.9)) / (2 * 1.5)
```

### Solution 2

```{r, title = "Solution"}
storm <- read.csv(
  'data/StormEvents.csv',
  stringsAsFactors = TRUE)
storm$STATE <- factor(storm$STATE)
```

### Solution 3

```{r, title = "Solution"}
x <- list(3, 4, 5, 7)
typeof(x)
```

The variable `x` has a data type of `list`, so R does not restrict its elements
to a particular type as it does for vectors. 

```{r, title = "Solution"}
typeof(c(income, x))
```

The result of combining a list and vector is a list, because the list is the
more flexible data structure.

### Solution 4

```{r, title = "Solution"}
species <- c('ape', 'bat', 'cat', 'dog')
abund <- 1:4
data <- data.frame(species, abund)
```

### Solution 5

```{r, title = "Solution"}
days[c(-1, -7)]
```

```{r, title = "Solution"}
days[seq(2, 7, 2)]
```

### Solution 6

```{r, title = "Solution"}
first <- function(dat) {
    if (is.vector(dat)) {
        result <- dat[[1]]
    } else if (is.matrix(dat)) {
        result <- dat[[1, 1]]
    } else {
        result <- dat[1, ]
    }
    return(result)
}
```

```{r, title = "Solution"}
m <- matrix(1:9, nrow = 3, ncol = 3)
first(m)
```
