---
---

## Load Data

We will use the function `read.csv()` to load data from a Comma Separated Value
file. The essential argument for the function to read in data is the path to the
file, other optional arguments adjust how the file is read.

Additional file types can be read in using `read.table()`; in fact, `read.csv()`
is a simple wrapper for the `read.table()` function having set some default
values for some of the optional arguments (e.g. `sep = ","`).
{:.notes}

===

Type `read.csv(` into the console and then press **tab** to see what arguments
this function takes. Hovering over each item in the list will show a description
of that argument from the help documentation about that function. Specify the
values to use for an argument using the syntax `name = value`.

```{r}
read.csv(
  file = 'data/species.csv',
  stringsAsFactors = FALSE)
```

===

Question
: How does `read.csv` determine the field names?

Answer
: {:.fragment} The `read.csv` command assumes the first row in the file contains
column names. Look at `?read.csv` to see the default `header = TRUE` argument.
What exactly that means is described down in the "Arguments" section.

===

Use the assignment operator "<-" to load data into a variable for
subsequent operations.

```{r handout = 0}
animals <- read.csv(file = 'data/animals.csv')
``` 

===

After reading in the "animals.csv" file, you can explore what types of data are
in each column with the `str` function, short for "structure".

```{r}
str(animals)
```

===

Missing data, as interpreted by the `read.csv` function, is controlled by the
`na.strings` argument. Override the default value of `'NA'` with the empty
string, `''`, to properly interpret the "species_id" and "sex" columns.

You can also specify multiple things to be interpreted as missing values, such
as `na.strings = c("missing", "no data", "< 0.05 mg/L", "XX")`.
{:.notes}

```{r handout = 0}
animals <- read.csv(file = 'data/animals.csv', na.strings = '')
``` 

```{r}
str(animals)
```
