---
---

## Distributions and Statistics

Since it was designed by statisticians, R can easily draw random numbers from
probability distributions and calculate probabilities.

===

To generate random numbers from a normal distribution, use the function
`rnorm()`

```{r handout = 0}
rnorm(n = 10)
```

===

| Function  | Returns                                      | Notes                     |
|-----------+----------------------------------------------+---------------------------|
| `rnorm()` | Draw random numbers from normal distribution | Specify `n`, `mean`, `sd` |
| `dnorm()` | Probability density at a given number        |                           |
| `pnorm()` | Cumulative probability up to a given number  | left-tailed by default    |
| `qnorm()` | The quantile given a cumulative probability  | opposite of pnorm         |

===

Statistical distributions and their functions.
See *Table 14.1* in **R for Everyone** by Jared Lander for a full table.

| Distribution | Functions     |
|--------------+---------------|
| Normal       | *norm         |
| Binomial     | *binom        |
| Poisson      | *pois         |
| Gamma        | *gamma        |
| Exponential  | *exp          |
| Uniform      | *unif         |
| Logistic     | *logis        |

===

R has built in functions for handling many statistical tests. 

```{r, include = FALSE}
set.seed(2345)
```
```{r handout = 0}
x <- rnorm(n = 100, mean = 15, sd = 7)
y <- rbinom(n = 100, size = 20, prob = .85)
```

===

The samples above are drawn from different distributions with different means.
The  T-Test should easily distinguish them, although it does not check
assumptions!

```{r}
t.test(x, y)
```

===

Shapiro's test of normality provides one routine for verifying assumptions.

```{r}
shapiro.test(y)
```
