---
---

## Data Structures

A data frame is a compound object, built up from (eventually) a few basic data
types, but there are intermediate objects to understand. Like all data frames,
`storm` is actually a "list".

```{r}
typeof(storm)
```

===

The "list" is one of three one-dimensional data structurs you will regularly
encounter.

- Lists
- Vectors
- Factors

===

## Lists

Lists are one-dimensional and each element is entirely unrestricted; you can put
anything in a list.

===

```{r}
x <- list('abc', 123, sin)
str(x)
```

===

Question
: Compare the structure of `storm` and `x` while thinking about the length of
each of their elements. Do the elements within list `x` have a lenght? The same
length?

Answer
: {:.fragment} The elements of `x` each has a length of 1. For example, the
command `length('abc')` yields `1`.

When you enter a single number or character string in R, you are actually
creating a one-dimensional data structure of length 1. There are not really
0-dimensional "scalars" in R. The kind of one-dimensional structure created in
this was is called a "vector".
{:.notes}

===

## Vectors

Vectors are an array of values of the same data type. Create a vector by
combining elements of the same type together using the function `c()`.

```{r handout = 0}
c(1, 2, 3)
```

===

All elements of an vector must be the same type, so when you attempt to combine
different types they will be coerced to the most flexible type.

```{r}
c(1, 2, 'c')
```

The difference between `c(1, 2, 3)` and `c(1, 2, 'c')` isn't just in the third
element. To understand the difference, we need to recognize data types.

===

## Data Types

A data frame has now been recognized as a list, and the `str` command gives an
indication that each field has a data type. The final key property is its
dimension.

```{r}
dim(storm)
```

A data frame is essentially a list of vectors, with an important constraint. The
vectors must all be of the same length, giving a rectangular data structure.
{:.notes}

===

Here is a summary of the data types you frequently encounter in data frames.

| Type      | Example           |
|-----------+-------------------|
| double    | 3.1, 4, Inf, NaN  |
| integer   | 4L, length(...)   |
| character | 'a', '4', 'ðŸ‘'    |
| logical   | TRUE, FALSE       |

Both the double and integer data types are considered numeric, and while
the `str` function tells you that a double is "num", the `typeof` function
will properly identify either numeric type. Missing data created with `NA`
actually have a variant for each data type. So you can put `NA` in any vector
without breaking the fule that the elements of a vector have the same data type.
{:.notes}

===

## Factors

A factor is a vector that can contain only predefined values, and is used to
store categorical data. Factors are like integer vectors, but posess a `levels`
attribute that assigns names to each level, or possible value in the vector.

===

Use `factor()` to create a vector with predefined values, which are often
character strings.

```{r handout = 0}
education <- factor(
    c('college', 'highschool', 'college', 'middle', 'middle'),
    levels = c('middle', 'highschool', 'college'))
```

===

The `str` function notes the labels, but prints the integers assigned in their
stead.

```{r}
str(education)
```

===

## Data Frames

One last property makes data frames stand out from lists: a data frame is a
list of *equal-length* vectors having *unique names*.

===

The "columns" visible in the data viewer are accessed by the `names` function.

```{r}
names(storm)
```

===

In summary, a data frame is the data structure most similar to a spreadsheet,
with a few key differences:

- The columns are **equal-length** vectors.
- As vectors, a column cannot hold values of the wrong type.
- Each column has a unique name.

===

Creating a data frame from scratch can be done by combining vectors with the
`data.frame()` function.

```{r handout = 0}
income <- c(32000, 28000, 89000, 0, 0)
df <- data.frame(education, income)
```

===

Remember to use these functions when getting to know a data frame:

| `dim()`            | dimensions              |
| `nrow()`, `ncol()` | number of rows, columns |
| `names()`          | (column) names          |
| `str()`            | structure               |
| `summary()`        | summary info            |
| `head()`           | shows beginning rows    |

===

## Matrices

One way to understand the need for another data structure, the matrix, is that a
matrix differs from a data frame in terms of the underlying data type.

===

| Dimensions | Homogeneous | Heterogeneous  |
|------------+-------------+----------------|
| 1-D        | `c()`       | `list()`       |
| 2-D        | `matrix()`  | `data.frame()` |
| n-D        | `array()`   |                |
