---
---

## The Console

The interpreter accepts commands interactively through the console.

===

Basic math, as you would type it on a calculator, is usually a valid command in the R language:

```{r}
1 + 2
```

```{r}
4^2
```

===

Question
: Why is the output prefixed by `[1]`?

Answer
: {:.fragment} That's the index, or position in a vector, of the first result.

===

A command giving a vector of results shows this clearly:

```{r}
seq(1, 100)
```

===

The interpreter understands more than arithmatic operations.
That last command told it to use (or "call") the function `seq()`.

Most of "learning R" involves getting to know a whole lot of functions, the effect of each function's  arguments (e.g. the input values `1` and `100`), and what each function returns (e.g. the output vector).
{:.notes}

===

## R as Calculator

A good place to begin learning R is with its built-in mathematical functionality.

===

## Arithmatic operators

Try `+`, `-`, `*`, `/`, and `^` (for raising to a power).

```{r}
5/3
```

===

## Logical tests

Test equality with `==` and inequality with `<=`, `<`, `!=`, `>`, or `>=`.

```{r}
1/2 == 0.5
```

===

## Math functions

Common mathematical functions like `sin`, `log`, and `sqrt`, exist along side some universal constants.

```{r}
sin(2 * pi)
```

===

## Programming idoms

Common computer programming functions like 'rep', 'sort', and 'range'

```{r}
rep(2, 5)
```

===

## Parentheses

Sandwiching something with `(` and `)` has two possible meanings.

===

Group sub-expressions by parentheses on an as-needed basis.

```{r}
(1 + 2) / 3
```

===

Call functions by typing their name and comma-separated arguments between parentheses.

```{r}
logb(2, 2)
```

===

## Assignment

When you start a new session, the R interpreter already recognizes many things, including

- any number
- any string of characters
- nearly universal operators (e.g. `+` and `/`)
- operators specific to R (e.g. `$` and `%*%`)
- functions in "base R"

===

To reference a number or function just type it in as above.
To referece a string of characters, surround them in quotation marks.

```{r}
'ab.cd'
```

===

Without quotation marks, the interpreter checks for things named `ab.cd` and
doesn't find anything:

```{r error = TRUE}
ab.cd
```

===

Question
: Is it better to use `'` or `"`?

Answer
: {:.fragment} Neither one is better. You will often encounter stylistic choices
like this, so if you don't have a personal preference try to mimic existing
styles.

===

You can expand the vocabulary known to the R interpreter by creating a new
**variable**. Using the symbol `<-` is referred to as assignment: the output of
any command to the right of `<-` gets the name given on its left.

```{r}
x <- seq(0, 100)
```

===

You'll notice that nothing prints to the console, because we assigned the output to a variable.
We can print the value of `x` by evaluating it without assignment.

```{r}
x
```

Assigning values to new variables (to the left of a `<-`) is the only time you
can reference something previously unknown to the interpreter. All other
commands must reference things already in the interpreter's vocabulary.
{:.notes}

===

Once assigned to a variable, a value becomes known to R and you can refer to it in other commands.

```{r}
plot(x, sin(x * 2 * pi / 100))
```
