---
---

## The Editor

The **console** is for evaluating commands you don't intend to keep or reuse. It's useful for testing commands and poking around.
The **editor** is where you compose scripts that will process data, perform analyses, code up visualizations, and even write reports.

These work together in RStudio, which has multiple ways to send parts of the script you are editing to the console for immediate evaluation.
Alternatively you can "source" the entire script.

===

Open up "{{ site.handouts }}" in the editor, and follow along by replacing the `...` placeholders with the code here. Then evalute just this line (Ctrl R on Windows, ⌘ R on Mac OS).

```{r title="{{ site.handouts }}"}
vals <- seq(1, 100)
```

===

Let's review the elements of this statement, from left to right:

- `vals` is the name of a (new) variable
- ` <- ` is an operator that assigns what's named on the left to equal the result of the expression on the right
- `seq` is the name of the function
- `(` is the opening paren of the function call
- `1` and `100` are both arguments, or parameters, to the function
- `)` is the closing paren of a function call

===

Question
: Why call `vals` a "variable" and `seq` a "function"?

Answer
: {:.fragment} It is true they are both names of objects known to R, and could be called variables. But `seq` has the important distinguishing feature of being **callable**, which is indicated in documentation by writing the function name with empty parens, as in `seq()`.

===

Our call to the function `seq` could have been much more explicit. We could give the arguments by the names that `seq` is expecting.

```{r title="{{ site.handouts }}"}
vals <- seq(from = 1,
            to = 100)
```

Run this code either line-by-line, or highlight the section to run (optionally with keyboard shortcut Ctrl-Return or ⌘ Return).

===

Question
: What's an advantage of naming arguments?

Answer
: {:.fragment} One advantage is that you can put them in any order. A related advantage is that you can then skip some arguments, which is fine to do if each skipped argument has a default value. A third advantage is code readability, which you should always be concious of while writing in the editor.

===

## Readability

Code readability in the editor cuts both ways: sometimes verbosity is useful, sometimes it is cumbersome.
The `seq()` function has an operator form available when only the `from` and `to` arguments are needed.

```{r}
1:100
```

The `:` operator should be used whenever possible because it replaces a common, cumbersome function call with an brief, intuitive syntax.
Likewise, the `assign` function duplicates the functionallity of the `<-` symbol, but is never used when the simpler operator will suffice.

===

## Function documentation

How would you get to know these properties and the names of a function's arguments?

```{r eval=FALSE}
?seq
```

How would you even know what function to call?

```{r eval=FALSE}
??sequence
```
