---
---

## Flow control

The R interpreter's "focus" flows through a script (or any section of code you run) line by line.
Without additional instruction, every line is processed from the top to bottom.
"Flow control" is the generic term for causing the interpreter to repeat or skip certain lines, using concepts like "for loops" and "if/else conditionals".

===

Flow control happens within blocks of code isolated between curly braces `{` and `}`, known as "statements".

```{r eval=FALSE, title="{{ site.handouts }}"}
if (...) {
    ...
} else {
    ...
}
```

The keyword `if` must be followed by a logical test which determines, at runtime, what to do next.
The R interpreter goes to the first statement if the logical value is `TRUE` and to the second statement if it's `FALSE`.

===

An if/else conditional would allow the `first` function to avoid the error thrown by calling `first(counts)`.

```{r title="{{ site.handouts }}"}
first <- function(dat) {
    if (is.vector(dat)) {
        result <- dat[1]
    } else {
        result <- dat[1, ]
    }
    return(result)
}
```

===

```{r}
first(df)
```

```{r}
first(counts)
```

<!--
===

## Exercise 7

The keywords `else` and `if` can be combined to allow flow control among more than two statements.

```{r eval=FALSE, title="{{ site.handouts }}"}
if (...) {
    ...
} else if {
    ...
} else if {
   ...
} else {
   ...
}
```

Expand the `first` function once again to differentiate between `dat` provided as a `matrix` and as a `data.frame`. It's up to you what the "first" element of a matrix should be!

[View solution](#solution-7)
{:.notes}
-->