---
---

## Load data into R

We will use the function `read.csv()` that reads a Comma-Separated-Values file.
The essential argument for the function to read in data is the path to the file,
and optional additional arguments that adjust how the file is interpreted.

Additional file types can be read in using `read.table()`; in fact, `read.csv()`
is a simple wrapper for the `read.table()` function having set some default
values for some of the optional arguments (e.g. `sep = ","`).
{:.notes}

===

Type `read.csv(` into the console and then press **tab** to see what arguments
this function takes. Hovering over each item in the list will show a description
of that argument from the help documentation about that function. Specify the
values to use for an argument using the syntax `name = value`.

```{r}
read.csv(file = "data/plots.csv", header = TRUE)
```

===

Question
: Is the `header` argument necessary?

Answer
: {:.fragment} No. Look at `?read.csv` to see that `TRUE` is the default value for this argument.

===

Use the assignment operator "<-" to store that data in memory and work with it

```{r title="{{ site.handouts[0] }}"}
plots <- read.csv(file = "data/plots.csv")
animals <- read.csv(file = "data/animals.csv")
``` 

You can specify what indicates missing data in the read.csv function using
either `na.strings = "NA"` or `na = "NA"`. You can also specify multiple things
to be interpreted as missing values, such as `na.strings = c("missing", "no
data", "< 0.05 mg/L", "XX")`.
{:.notes}

===

After reading in the "animals.csv" and "plots.csv" files, let's explore what
types of data are in each column and what kind of structure your data has.

```{r}
str(plots)
```

```{r}
summary(plots)
```
