---
---

## Distributions and Statistics

Since it is designed for statistics, R can easily draw random numbers from statistical distributions and calculate distribution values. 

To generate random numbers from a normal distribution, use the function `rnorm()`

```{r}
ten_random_values <- rnorm(n = 10)
```

===

| Function  | Returns                                      | Notes                     |
|-----------+----------------------------------------------+---------------------------|
| `rnorm()` | Draw random numbers from normal distribution | Specify `n`, `mean`, `sd` |
| `dnorm()` | Probability density at a given number        |                           |
| `pnorm()` | Cumulative probability up to a given number  | left-tailed by default    |
| `qnorm()` | The quantile given a cumulative probability  | opposite of pnorm         |

===

Statistical distributions and their functions.
See *Table 14.1* in **R for Everyone** by Jared Lander for a full table.

| Distribution | Random Number |
|--------------+---------------|
| Normal       | rnorm         |
| Binomial     | rbinom        |
| Poisson      | rpois         |
| Gamma        | rgamma        |
| Exponential  | rexp          |
| Uniform      | runif         |
| Logistic     | rlogis        |

===

R has built in functions for handling many statistical tests. 

```{r title="{{ site.handouts }}"}
x <- rnorm(n = 100, mean = 25, sd = 7)
y <- rbinom(n = 100, size = 50, prob = .85)
```

```{r}
t.test(x, y)
```

===

Linear regression with the `lm()` function uses a formula notation to specify relationships between variables (e.g. `y ~ x`).

```{r title="{{ site.handouts }}"}
fit <- lm(y ~ x)
```

```{r}
summary(fit)
```

===

## Exercise 4

Create a data frame from scratch that has three columns and 5 rows. In column "size" place a sequence from 1 to 5. For column "year", create a factor with three levels representing the past three years. In column "prop", place 5 random samples from a uniform distribution. Show the summary of a linear model following the formula "prop ~ size + year".
