---
---

## Data Types

A data frame is clearly a table, but what exactly is a table in the R
environment? The `str` command gave an indication that each field has it's own
data type.

===

| Type      | Example           |
|-----------+-------------------|
| double    | 3.1, 4, Inf, NaN  |
| integer   | 4L, 0L, 999L      |
| character | 'a', '4', '👏'    |
| logical   | TRUE, FALSE       |
| missing   | NA                |

===

## Data structures

A data frame is a compound object, built from one or more objects that hew to
the basic data types. Like all data frames, "animals" is a "list".

```{r}
is.list(animals)
```


===

The "list" is one of three one-dimensional data structurs you will regularly
encounter.

- Lists
- Factors
- Vectors

===

## Vectors

Vectors are an array of values of the same type. Create a vector by combining
elements of the same type together using the function `c()`.

```{r title="{{ site.handouts[0] }}"}
counts <- c(4, 3, 7, 5, 2)
```

===

All elements of an vector must be the same type, so when you attempt to combine
different types they will be coerced to the most flexible type.

```{r}
c(1, 2, 'c')
```

===

## Factors

A factor is a vector that can contain only predefined values, and is used to
store categorical data. Factors are like integer vectors, but posess a `levels`
attribute that assigns names to however many discrete categories are specified.

===

Use `factor()` to create a vector with predefined values, which are often
characters or "strings".

```{r title="{{ site.handouts[0] }}"}
education <- factor(
    c('college', 'highschool', 'college', 'middle', 'middle'),
    levels = c('middle', 'highschool', 'college'))
```

===

The `str` function notes the labels, but prints the integers assigned in their
stead.

```{r}
str(education)
```

===

## Lists

Lists are like vectors and factors, but their elements can be of any data type
or structure.

===

Construct lists with `list()` instead of `c()`:

```{r}
list(1, 2, 'c')
```

===

Lists can include vectors, factors, and even other lists.

```{r}
list(1, c('a', 'b'))
```

===

Now we can anwser the question, "What is a table in the R environment?": it is a
list of *equal-length* vectors having unique names.

===

## Data frames

This is the data structure most similar to a spreadsheet, with a few key
differences:

- The columns are **equal-length** vectors.
- As vectors, a column cannot hold values of the wrong type.
- The columns have a unique name, along with their position or order.

===

Creating a data frame from scratch can be done by combining vectors with the
`data.frame()` function.

```{r title="{{ site.handouts[0] }}"}
df <- data.frame(education, counts)
```

===

There are several functions to get to know a data frame:

| `dim()`            | dimensions              |
| `nrow()`, `ncol()` | number of rows, columns |
| `names()`          | (column) names          |
| `str()`            | structure               |
| `summary()`        | summary info            |
| `head()`           | shows beginning rows    |

===

```{r}
names(df)
```

===

## Tables, Matrices & Arrays

One way to understand the need for different data structures, is that they serve
the purpose of holding data with different sorts of complexity.

===

| Dimensions | Homogeneous | Heterogeneous  |
|------------+-------------+----------------|
| 1-D        | `c()`       | `list()`       |
| 2-D        | `matrix()`  | `data.frame()` |
| n-D        | `array()`   |                |
